Clustering based on probability distributions with application on residential customers

### 2021-10-19 main.Rmd
simulation till 20oct

To validate the clustering approaches, we create data designs that replicate prototype behaviors that might be seen in electricity data contexts. We spiked several features in the data to see where one method works better than the other and where they might give us the same outcome or the effect of missing data <!---and trends--> on the proposed methods. A continuous measured variable $y$ of length $T$ indexed by $\{0, 1, \dots T-1\}$. Three circular granularities $g1$, $g2$ and $g3$ are considered with $2$, $3$ and $5$ levels respectively. Categories of $g1$, $g2$ and $g3$ are represented by ${g10,g11}$, ${g20, g21, g22}$ and ${g30, g31, g32, g33, g34}$. These categories could be integers or some more meaningful labels. For example, the granularity "day-of-week" could be either represented by ${0, 1, 2, \dots, 6}$ or ${Mon, Tue, \dots, Sun}$. We created independent replications $R = \{25, 250, 500\}$ of all data designs to see if our proposed clustering approaches can detect distinct designs in various groups for small, medium and large number of series. A sample size of $T=\{300, 1000, 5000\}$ is used in all designs to test small, medium and large sized series. The method could preform differently with different jumps between consecutive categories. So a mean difference of $diff = \{1, 2, 5\}$ for corresponding categories are also considered. The code for creating these designs can be found in the Supplementary section (link to github repo). The results for $T=300$ and $R=25$ is shown, that means we have $25$ time series each with length $300$. The rest of the results could be found in the supplementary paper.

@@

_All signifiant granularities_

@@

Consider the scenario when all three granularities $g1$, $g2$, and $g3$ are responsible for distinguishing the designs. That means that for at least one among the to-be-grouped designs, the pattern for each of $g1$, $g2$, and $g3$ will change. We consider different distributions across categories (as in Table \ref{tab:tab-dist-design}) (top) that will lead to different designs (as in Table \ref{tab:tab-dist-design}) (below). Figure \ref{fig:plot-3gran-new} shows the linear and cyclic representation of the simulated variable under these five designs. For the consequent data designs, only graphical representations will be provided.

```{r tab-distribution}
granularity = c("g1", "g2", "g3")
# alternate_design =  c("g10 ~ N(0, 1), g11 ~ N(2, 1)",
#                       "g21 ~ N(2, 1), g22 ~ N(1, 1), g23 ~ N(0, 1)",
#                       "g31 ~ N(0, 1), g32 ~ N(1, 1), g33 ~ N(2, 1), g34 ~ N(1, 1), g35 ~ N(0, 1)")

@@

_Few significant granularities_

This is the case where one granularity will remain the same across all designs.
We consider the case where the distribution of $y$ would vary across levels of $g2$ for all designs, across levels of $g1$ for few designs and $g3$ does not change across designs. So $g3$ is not responsible for distinguishing across designs. Figure \ref{#fig:plot-linear} shows the linear and cyclic representation of $y$. The first panel shows raw plot of $y$ in a linear scale and the second panel shows distribution of $y$ across cyclic granularities namely $g1$, $g2$ and $g3$. As could be seen from the plots, it is impossible to decipher from the raw time plot that the time series variable shows such pattern across different granularities.

_One signifiant granularity_

Here only one granularity is responsible for distinguishing the designs. Designs change significantly only for the granularity $g2$. Figure \ref{fig:plot-1gran} shows this.

@@

A subset of many possible designs are shown in Figure \ref{fig: plot-linear }. For the parameter space (XXX unique  combinations shown in table YYY), 100, 500 independent replications of all possible combination of simulation parameters were generated. The clustering methodologies were run all these unique combinations and subsets of these to verify if the methodologies work as expected.

@@

p_null <- sim_panel_null %>%
  ggplot(aes(x = as.factor(g2), y = sim_data)) +
  facet_wrap(~g1, labeller = "label_both") +
  geom_boxplot() +
  ylab("") +
  #ggtitle(paste("(a)", round(null, 2))) +
  xlab("") +
  theme_bw()+ stat_summary(
    fun = median,
    geom = 'line',
    aes(group = 1), size = 0.8, color = "blue") +
  #ggtitle("design1")+
  theme_validation() +ylab("")  +
  theme(panel.spacing =unit(0, "lines"))+ theme(
    strip.text = element_text(size = 8, margin = margin(b = 0, t = 0)))+
  theme(plot.margin = margin(0, 0, 0, 0, "cm") )

p_varf <- sim_panel_varf %>%
  ggplot(aes(x = as.factor(g2), y = sim_data)) +
  facet_wrap(~g1, labeller = "label_both") +
  geom_boxplot() +
  ylab("") +
  #ggtitle(paste("(b)", round(varf, 2))) +
  xlab("")+
  theme_bw()+ stat_summary(
    fun = median,
    geom = 'line',
    aes(group = 1), size = 0.8, color = "blue") +
  #ggtitle("design2")+
  theme_validation()+ylab("") +
  theme(panel.spacing =unit(0, "lines"))+ theme(
    strip.text = element_text(size = 8, margin = margin(b = 0, t = 0)))+
  theme(plot.margin = margin(0, 0, 0, 0, "cm") )

p_varx <- sim_panel_varx %>%
  ggplot(aes(x = as.factor(g2), y = sim_data)) +
  facet_wrap(~g1, labeller = "label_both") +
  geom_boxplot()+
  ylab("")  +
  #ggtitle(paste("(c)", round(varx, 2))) +
  xlab("")+
  theme_bw()+ stat_summary(
    fun = median,
    geom = 'line',
    aes(group = 1), size = 0.8, color = "blue") +
  #ggtitle("design3")+
  theme_validation()+ylab("") +
  theme(panel.spacing =unit(0, "lines"))+ theme(
    strip.text = element_text(size = 8, margin = margin(b = 0, t = 0)))+
  theme(plot.margin = margin(0, 0, 0, 0, "cm") )

p_varall <- sim_panel_varall %>%
  ggplot(aes(x = as.factor(g2), y = sim_data)) +
  facet_wrap(~g1, labeller = "label_both") +
  geom_boxplot() +
  ylab("") +
  #ggtitle(paste("(d)", round(varall, 2))) +
  xlab("")+
  theme_bw()+ stat_summary(
    fun = median,
    geom = 'line',
    aes(group = 1), size = 0.8, color = "blue") +
  #ggtitle("design4") +
  theme_validation()+ylab("") +
  theme(panel.spacing =unit(0, "lines"))+ theme(
    strip.text = element_text(size = 8, margin = margin(b = 0, t = 0)))+
  theme(plot.margin = margin(0, 0, 0, 0, "cm") )

@@

#((p1 + p_null)/( p2 + p_varf)/(p3 + p_varx)/(p4 + p_varall)) 
#p_null+ p_varf + p_varx + p_varall
ggpubr::ggarrange(p_null, p_varf, p_varx, p_varall, labels = c("Design-1", "Design-2", "Design-3", "Design-4"))

@@

| Granularity type                                                   	| # Significant 	| # Replications 	|
|--------------------------------------------------------------------	|---------------	|----------------	|
| **Individual**  <br><br># obs: 300, 500, 2000  <br># clusters: 6/7 	| 1/2/3         	| 25, 100, 200   	|
| **Interaction**  <br><br># obs: 500, 2000  <br># clusters: 4       	| 1/2           	| 25, 100, 200   	|

>>>>

To validate the clustering approaches, we create data designs that replicate prototype behaviors that might be seen in electricity data contexts. We spiked several attributes in the data to see where one method works better than the other and where they might give us the same outcome or the effect of missing data <!---and trends--> on the proposed methods. A continuous measured variable $y$ of length $T$ indexed by $\{0, 1, \dots T-1\}$. Three circular granularities $g1$, $g2$ and $g3$ are considered with $2$, $3$ and $5$ levels respectively. Categories of $g1$, $g2$ and $g3$ are represented by ${g10,g11}$, ${g20, g21, g22}$ and ${g30, g31, g32, g33, g34}$. These categories could be integers or some more meaningful labels. For example, the granularity "day-of-week" could be either represented by ${0, 1, 2, \dots, 6}$ or ${Mon, Tue, \dots, Sun}$. We created independent replications $R = \{25, 250, 500\}$ of all data designs to see if our proposed clustering approaches can detect distinct designs in various groups for small, medium and large number of series. A sample size of $T=\{300, 1000, 5000\}$ is used in all designs to test small, medium and large sized series. The methods could preform differently with different jumps between consecutive categories. So a mean difference of $diff = \{1, 2, 5\}$ for corresponding categories are also considered. The code for creating these designs can be found in the Supplementary section (link to github repo). The results for $T=300$ and $R=25$ is shown, that means we have $25$ time series each with length $300$. The rest of the results could be found in the supplementary paper.

@@

  _(a) All signifiant granularities_

@@

\noindent Consider the scenario when all three granularities $g1$, $g2$, and $g3$ are responsible for distinguishing the designs. That means that for at least one among the to-be-grouped designs, the pattern for each of $g1$, $g2$, and $g3$ will change. We consider different distributions across categories (as in Table \ref{tab:tab-dist-design}) (top) that will lead to different designs (as in Table \ref{tab:tab-dist-design}) (below). Figure \ref{fig:plot-3gran-new} shows the linear and cyclic representation of the simulated variable under these five designs. For the consequent data designs, only graphical representations will be provided.

_(b) Few significant granularities_

\noindent This is the case where one granularity will remain the same across all designs. We consider the case where the distribution of $y$ would vary across levels of $g2$ for all designs, across levels of $g1$ for few designs and $g3$ does not change across designs. So $g3$ is not responsible for distinguishing across designs. Figure \ref{#fig:plot-linear} shows the linear and cyclic representation of $y$. The first panel shows raw plot of $y$ in a linear scale and the second panel shows distribution of $y$ across cyclic granularities namely $g1$, $g2$ and $g3$. As could be seen from the plots, it is impossible to decipher from the raw time plot that the time series variable shows such pattern across different granularities.

_(c) One signifiant granularity_

\noindent Here only one granularity is responsible for distinguishing the designs. Designs change significantly only for the granularity $g2$. Figure \ref{fig:plot-1gran} shows this.

```{r tab-distribution, fig.pos="htb"}
granularity = c("g1", "g2", "g3")
# alternate_design =  c("g10 ~ N(0, 1), g11 ~ N(2, 1)",
#                       "g21 ~ N(2, 1), g22 ~ N(1, 1), g23 ~ N(0, 1)",
#                       "g31 ~ N(0, 1), g32 ~ N(1, 1), g33 ~ N(2, 1), g34 ~ N(1, 1), g35 ~ N(0, 1)")

@@

@@

<!-- A subset of many possible designs are shown in Figure \ref{fig: plot-linear }. For the parameter space (XXX unique  combinations shown in table YYY), 100, 500 independent replications of all possible combination of simulation parameters were generated. The clustering methodologies were run all these unique combinations and subsets of these to verify if the methodologies work as expected. -->

@@

p_null <- sim_panel_null %>%
  ggplot(aes(x = as.factor(g1), y = sim_data)) +
  facet_wrap(~g2, labeller = "label_both") +
  geom_boxplot() +
  ylab("") +
  #ggtitle(paste("(a)", round(null, 2))) +
  xlab("") +
  theme_bw()+ stat_summary(
    fun = median,
    geom = 'line',
    aes(group = 1), size = 0.8, color = "blue") +
  #ggtitle("design1")+
  theme_validation() +ylab("")  +
  theme(panel.spacing =unit(0, "lines"))+ theme(
    strip.text = element_text(size = 8, margin = margin(b = 0, t = 0)))+
  theme(plot.margin = margin(0, 0, 0, 0, "cm") ) +ggtitle("Design-1")

p_varf <- sim_panel_varf %>%
  ggplot(aes(x = as.factor(g1), y = sim_data)) +
  facet_wrap(~g2, labeller = "label_both") +
  geom_boxplot() +
  ylab("") +
  #ggtitle(paste("(b)", round(varf, 2))) +
  xlab("")+
  theme_bw()+ stat_summary(
    fun = median,
    geom = 'line',
    aes(group = 1), size = 0.8, color = "blue") +
  #ggtitle("design2")+
  theme_validation()+ylab("") +
  theme(panel.spacing =unit(0, "lines"))+ theme(
    strip.text = element_text(size = 8, margin = margin(b = 0, t = 0)))+
  theme(plot.margin = margin(0, 0, 0, 0, "cm") )+ggtitle("Design-2")

p_varx <- sim_panel_varx %>%
  ggplot(aes(x = as.factor(g1), y = sim_data)) +
  facet_wrap(~g2, labeller = "label_both") +
  geom_boxplot()+
  ylab("")  +
  #ggtitle(paste("(c)", round(varx, 2))) +
  xlab("")+
  theme_bw()+ stat_summary(
    fun = median,
    geom = 'line',
    aes(group = 1), size = 0.8, color = "blue") +
  #ggtitle("design3")+
  theme_validation()+ylab("") +
  theme(panel.spacing =unit(0, "lines"))+ theme(
    strip.text = element_text(size = 8, margin = margin(b = 0, t = 0)))+
  theme(plot.margin = margin(0, 0, 0, 0, "cm") )+ggtitle("Design-3")

p_varall <- sim_panel_varall %>%
  ggplot(aes(x = as.factor(g1), y = sim_data)) +
  facet_wrap(~g2, labeller = "label_both") +
  geom_boxplot() +
  ylab("") +
  #ggtitle(paste("(d)", round(varall, 2))) +
  xlab("")+
  theme_bw()+ stat_summary(
    fun = median,
    geom = 'line',
    aes(group = 1), size = 0.8, color = "blue") +
  #ggtitle("design4") +
  theme_validation()+ylab("") +
  theme(panel.spacing =unit(0, "lines"))+ theme(
    strip.text = element_text(size = 8, margin = margin(b = 0, t = 0)))+
  theme(plot.margin = margin(0, 0, 0, 0, "cm") )+ggtitle("Design-4")

@@

#((p1 + p_null)/( p2 + p_varf)/(p3 + p_varx)/(p4 + p_varall)) 
#p_null+ p_varf + p_varx + p_varall
ggpubr::ggarrange(p_null, p_varf, p_varx, p_varall)

#labels = c("Design-1", "Design-2", "Design-3", "Design-4"))

@@

<!-- | Granularity type                                                   	| # Significant 	| # Replications 	| -->
<!-- |--------------------------------------------------------------------	|---------------	|----------------	| -->
<!-- | **Individual**  <br><br># obs: 300, 500, 2000  <br># clusters: 4/5 	| 1/2/3         	| 25, 100, 200   	| -->
<!-- | **Interaction**  <br><br># obs: 500, 2000  <br># clusters: 4       	| 2           	| 25, 100, 200   	| -->
